---
title: "REU_R_2018_workshop_2"
author: "Ilya"
date: "6/20/2018"
output: github_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#####packages
```{r}
#add dependencies = TRUE to make sure R installs packages needed by this package
#install package
install.packages("ggplot2", dependencies = TRUE)
install.packages("dplyr", dependencies = TRUE)
#load package so you can use it here
library(ggplot2)
library(dplyr)
```


#####set working directory
```{r}
setwd("~/ilya documents/R/REU/workshops 2018/workshop_2/REU_R_2018_workshop_2")
```

#####read in PanTHERIA data (source: http://esapubs.org/archive/ecol/e090/184/#data)
#####open in excel and save as csv
#####and inspect first few and last few lines 
```{r}
#read in; row.names must be NULL or otherwise get error related to duplicate row names
P <- read.csv("PanTHERIA_1-0_WR05_Aug2008.csv", header=TRUE)
#note: reading in as .txt resulted in Genus and species getting split up
#P <- read.table("PanTHERIA_1-0_WR05_Aug2008.txt", header=TRUE, row.names=6)

#view first few lines
head(P)
#view last few lines
tail(P)
```

#####get the dimensions of dataset
```{r}
dim(P)
```

#####print list of column names of dataset
```{r}
names(P)
```

#####get a summary of your data
```{r}
summary(P)
```

####summarize single column
```{r}
# use the $ operator to access a single column
summary(P$X5.1_AdultBodyMass_g)
#notice there are -999 values; need to change to NA
```

#####change -999 to NA in single column using function "which"
```{r}
#see what == does
TRUE==TRUE
1==2
#see what != does
TRUE!=FALSE
1!=1
#find -999 rows
row_indices = which(P$X5.1_AdultBodyMass_g==-999)
#change values to NA
P$X5.1_AdultBodyMass_g[row_indices]=NA
#get summary again
summary(P$X5.1_AdultBodyMass_g)
```

#####change -999 to NA in all columns using function replace
```{r}
#check that there are -999 in various cols
summary(P)
P = replace(P, P==-999, NA)
#check that now there are NAs, no longer -999 as min
summary(P)
```



#####get more summary stats on single column
#####add argument na.rm =TRUE to get summary of non-NAs 
```{r}
# run various summary stats column length
length(P$X5.1_AdultBodyMass_g)
# minimum value in column
min(P$X5.1_AdultBodyMass_g, na.rm=TRUE)
# maximum value in column
max(P$X5.1_AdultBodyMass_g, na.rm=TRUE)
# range of values in column, min-max
range(P$X5.1_AdultBodyMass_g, na.rm=TRUE)
# mean of column, removes NAs
mean(P$X5.1_AdultBodyMass_g, na.rm=TRUE)
# standard deviation of column
sd(P$X5.1_AdultBodyMass_g, na.rm=TRUE)
```

#####Indexing to get only rows and columns meeting criteria
```{r}
#mydataset[row(s) that meet this criteria, columns that meet this criteria]
P_index<-P[P$MSW05_Family=="Canidae",c(1:8)]
#inspect/summarize
head(P_index)
```

#####Indexing second example -- find out dog and cat species (no data about them)
```{r}
#mydataset[row(s) that meet this criteria, columns that meet this criteria]
P_index<-P[P$MSW05_Family %in% c("Canidae","Felidae"), c(1:6)]
#inspect/summarize: check that both dogs and cats are there using function unique
unique(P_index$MSW05_Family)
```
#####Subsetting -- get only dog records, first eight columns
```{r}
#subset(mydataset,row(s) that meet this criteria, select=c(column names or numbers))
P_subset<-subset(P,MSW05_Family=="Canidae",select=c(1:8))
#inspect and summarize
head(P_subset)
```

#more indexing and subsetting examples
```{r}
#get data for nocturnal carnivores, including all columns
P_carn_noct_index = P[P$X1.1_ActivityCycle==1 & P$MSW05_Order=="Carnivora",]
#OR
P_carn_noct_subset = subset(P,X1.1_ActivityCycle==1 & MSW05_Order=="Carnivora")
#get data with AdultBodyMass_g greater than median, first 8 columns
median(P$X5.1_AdultBodyMass_g)
#indexing
P_himass_index = P[P$X5.1_AdultBodyMass_g> median(P$X5.1_AdultBodyMass_g),c(1:8)]
#subsetting
P_himass_subset = subset(P,X5.1_AdultBodyMass_g> median(P$X5.1_AdultBodyMass_g),select=c(1:8))
#check this is about half the data
dim(P_himass_subset)
#check both methods get same result
dim(P_himass_index)==dim(P_himass_subset)
```

####Data visualization
#####Histogram in base R
```{r}
hist(P$X15.1_LitterSize)
```

#####Histogram in ggplot2
```{r}
ggplot(data = P, mapping = aes(x = P$X15.1_LitterSize)) + 
  geom_histogram()
```

######Boxplot in base R
```{r}
boxplot(P$X15.1_LitterSize)
```

######Boxplot in ggplot2
```{r}
#set x = "" because there's just one variable
ggplot(data = P, mapping = aes(x="", y = X15.1_LitterSize))+
  geom_boxplot()
```

######Boxplot in ggplot2 for multiple variables
```{r}
#set x = Order
ggplot(data = P, mapping = aes(x=MSW05_Order, y = X15.1_LitterSize))+
  geom_boxplot()
```

######Boxplot in ggplot2 for multiple variables w/ legible axis labels
```{r}
#google search for : "vertical x axis labels ggplot2"
#set x = Order
ggplot(data = P, mapping = aes(x=MSW05_Order, y = X15.1_LitterSize))+
  geom_boxplot()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

#####Subset the data for Primates and Rodentia 
```{r}
Primates = subset(P, MSW05_Order == "Primates")
Rodentia = subset(P, MSW05_Order == "Rodentia")
#or
Primates = P[P$MSW05_Order== "Primates",]
Rodentia = P[P$MSW05_Order== "Rodentia",]
```

#####Report the number of species within Primates and Rodentia
```{r}
#use function dim, which returns two numbers -- rows and columns 
dim(Primates)
dim(Rodentia)
```

#####Report the mean and standard deviation of weaning age for Primates and Rodentia
```{r}
#primates
mean(Primates$X25.1_WeaningAge_d, na.rm=TRUE)
sd(Primates$X25.1_WeaningAge_d, na.rm=TRUE)
#rodents
mean(Rodentia$X25.1_WeaningAge_d, na.rm=TRUE)
sd(Rodentia$X25.1_WeaningAge_d, na.rm=TRUE)
```

#####Create a histogram and boxplot of social group size (X10.2_SocialGrpSize) for Primates
```{r}
#create plot with 1 row, 2 columns of subplots
par(mfrow=c(1,2))
hist(Primates$X25.1_WeaningAge_d)
boxplot(Primates$X25.1_WeaningAge_d)
```

#####Which Order, Primates or Rodentia, has more outliers for InterBirthInterval_d? Do outliers tend to have shorter or longer interbirth intervals? 
```{r}
#Primate outliers; use sort function to put them in order
primate_outliers <- boxplot.stats(Primates$X14.1_InterbirthInterval_d)$out
sort(primate_outliers)
#compare to average
summary(Primates$X14.1_InterbirthInterval_d)
#rodent outliers
rodent_outliers<-  boxplot.stats(Rodentia$X14.1_InterbirthInterval_d)$out
sort(rodent_outliers)
#compare to average
summary(Rodentia$X14.1_InterbirthInterval_d)
```

####Data manipulation with package dplyr
#####grouping and summarizing
```{r}
#1. Group species by Order and Family
by_Order_Family <- group_by(P, MSW05_Order, MSW05_Family)
#2. Summarise to compute means of variables of interest and number of species
by_Ord_Fam_sum <- summarise(by_Order_Family,
  count = n(),
  hr = mean(X22.2_HomeRange_Indiv_km2, na.rm = TRUE),
  mass = mean(X5.1_AdultBodyMass_g, na.rm = TRUE)
)
head(by_Ord_Fam_sum)#look at first 6 rows
```

#####grouping and summarizing using pipe operator
```{r}
#create by_Order_Family to hold summarized info
by_Order_Family_sum <-P %>%
#1. Group species by Order and Family
  group_by(MSW05_Order, MSW05_Family) %>%
#2. Summarise to compute means of variables of interest and number of species
    summarise(
    count = n(),
    hr = mean(X22.2_HomeRange_Indiv_km2, na.rm = TRUE),
    mass = mean(X5.1_AdultBodyMass_g, na.rm = TRUE)
  )
#make a plot of mass vs. hr size with size of dots proportional to number of species
ggplot(data = by_Ord_Fam_sum, mapping = aes(x = mass, y = hr)) +
  geom_point(aes(size = count), alpha = 1/3) +
  geom_smooth(se = FALSE)+
  scale_y_log10()+
  scale_x_log10()
```
