---
title: "REU_R_2018_workshop_2"
author: "Ilya"
date: "6/20/2018"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


#####packages
```{r}
#specify the repository (repos) to avoid error about mirror not being set
#also add logical so packages are only installed if they are missing 
#https://stackoverflow.com/questions/33969024/install-packages-fails-in-knitr-document-trying-to-use-cran-without-setting-a
if(!require(ggplot2)) install.packages("ggplot2",repos = "http://cran.us.r-project.org")
if(!require(dplyr)) install.packages("dplyr",repos = "http://cran.us.r-project.org")
if(!require(data.table)) install.packages("data.table",repos = "http://cran.us.r-project.org")
#load package so you can use it here
library(dplyr)
library(ggplot2)
library(data.table)
```


#####set working directory -- note you'll need to change the working directory to your own computer 
```{r}
setwd("~/ilya documents/R/REU/workshops 2018/workshop_2/REU_R_2018_workshop_2")
```

#####read in PanTHERIA data (source: http://esapubs.org/archive/ecol/e090/184/#data) and inspect first few and last few lines 
```{r}
#open txt file in excel, save as csv 
P <- read.csv("PanTHERIA_1-0_WR05_Aug2008.csv", header=TRUE)

#view first few lines
head(P)
#view last few lines
tail(P)
```

#####alternative way of reading in data: this is a more reliable method than opening .txt file in excel and saving as csv (as above with read.csv), because some .txt files will be too big to open in excel. however, remembered about this method after powerpoint was done, and this method results in variables having slightly different names vs. method above, which would be confusing, so not using for this workshop.   
```{r}
#use fread from package data.table
P_alternate <- fread("PanTHERIA_1-0_WR05_Aug2008.txt")

```


#####get the dimensions of dataset-- should be 5416 row*55 col
```{r}
dim(P)
```

#####print list of column names of dataset
```{r}
names(P)
```

#####get a summary of your data
```{r}
summary(P)
```

####summarize single column
```{r}
# use the $ operator to access a single column
summary(P$X5.1_AdultBodyMass_g)
#notice there are -999 values; need to change to NA
```

#####change -999 to NA in single column using function "which"
```{r}
#see what == does
TRUE==TRUE
1==2
#see what != does
TRUE!=FALSE
1!=1
#find -999 rows
row_indices = which(P$X5.1_AdultBodyMass_g==-999)
#change values to NA
P$X5.1_AdultBodyMass_g[row_indices]=NA
#get summary again
summary(P$X5.1_AdultBodyMass_g)
```

#####change -999 to NA in all columns using function replace
```{r}
#check that there are -999 in various cols
summary(P)
P = replace(P, P==-999, NA)
#check that now there are NAs, no longer -999 as min
summary(P)
```



#####get more summary stats on single column
#####add argument na.rm =TRUE to get summary of non-NAs 
```{r}
# run various summary stats column length
length(P$X5.1_AdultBodyMass_g)
# minimum value in column
min(P$X5.1_AdultBodyMass_g, na.rm=TRUE)
# maximum value in column
max(P$X5.1_AdultBodyMass_g, na.rm=TRUE)
# range of values in column, min-max
range(P$X5.1_AdultBodyMass_g, na.rm=TRUE)
# mean of column, removes NAs
mean(P$X5.1_AdultBodyMass_g, na.rm=TRUE)
# standard deviation of column
sd(P$X5.1_AdultBodyMass_g, na.rm=TRUE)
```

#####Indexing to get only rows and columns meeting criteria
```{r}
#mydataset[row(s) that meet this criteria, columns that meet this criteria]
P_index<-P[P$MSW05_Family=="Canidae",c(1:8)]
#inspect/summarize
head(P_index)
```

#####Indexing second example -- find out dog and cat species (no data about them)
```{r}
#mydataset[row(s) that meet this criteria, columns that meet this criteria]
P_index<-P[P$MSW05_Family %in% c("Canidae","Felidae"), c(1:6)]
#inspect/summarize: check that both dogs and cats are there using function unique
unique(P_index$MSW05_Family)
```
#####Subsetting -- get only dog records, first eight columns
```{r}
#subset(mydataset,row(s) that meet this criteria, select=c(column names or numbers))
P_subset<-subset(P,MSW05_Family=="Canidae",select=c(1:8))
#inspect and summarize
head(P_subset)
```

#####more indexing and subsetting examples
```{r}
#get data for nocturnal carnivores, including all columns
P_carn_noct_index = P[P$X1.1_ActivityCycle==1 & P$MSW05_Order=="Carnivora",]
#OR
P_carn_noct_subset = subset(P,X1.1_ActivityCycle==1 & MSW05_Order=="Carnivora")
#get data with AdultBodyMass_g greater than median, first 8 columns
median(P$X5.1_AdultBodyMass_g)
#indexing
P_himass_index = P[P$X5.1_AdultBodyMass_g> median(P$X5.1_AdultBodyMass_g),c(1:8)]
#subsetting
P_himass_subset = subset(P,X5.1_AdultBodyMass_g> median(P$X5.1_AdultBodyMass_g),select=c(1:8))
#check this is about half the data
dim(P_himass_subset)
#check both methods get same result
dim(P_himass_index)==dim(P_himass_subset)
```

####Data visualization
#####Histogram in base R
```{r}
hist(P$X15.1_LitterSize)
```

#####Histogram in ggplot2
```{r}
ggplot(data = P, mapping = aes(x = P$X15.1_LitterSize)) + 
  geom_histogram()
```

######Boxplot in base R
```{r}
boxplot(P$X15.1_LitterSize)
```

######Boxplot in ggplot2
```{r}
#set x = "" because there's just one variable
ggplot(data = P, mapping = aes(x="", y = X15.1_LitterSize))+
  geom_boxplot()
```

######Boxplot in ggplot2 for multiple variables
```{r}
#set x = Order
ggplot(data = P, mapping = aes(x=MSW05_Order, y = X15.1_LitterSize))+
  geom_boxplot()
```

######Boxplot in ggplot2 for multiple variables w/ legible axis labels
```{r}
#google search for : "vertical x axis labels ggplot2"
#set x = Order
ggplot(data = P, mapping = aes(x=MSW05_Order, y = X15.1_LitterSize))+
  geom_boxplot()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

#####Subset the data for Primates and Rodentia 
```{r}
Primates = subset(P, MSW05_Order == "Primates")
Rodentia = subset(P, MSW05_Order == "Rodentia")
#or
Primates = P[P$MSW05_Order== "Primates",]
Rodentia = P[P$MSW05_Order== "Rodentia",]
```

#####Report the number of species within Primates and Rodentia
```{r}
#use function dim, which returns two numbers -- rows and columns 
dim(Primates)
dim(Rodentia)
```

#####Report the mean and standard deviation of weaning age for Primates and Rodentia
```{r}
#primates
mean(Primates$X25.1_WeaningAge_d, na.rm=TRUE)
sd(Primates$X25.1_WeaningAge_d, na.rm=TRUE)
#rodents
mean(Rodentia$X25.1_WeaningAge_d, na.rm=TRUE)
sd(Rodentia$X25.1_WeaningAge_d, na.rm=TRUE)
```

#####Create a histogram and boxplot of social group size (X10.2_SocialGrpSize) for Primates
```{r}
#create plot with 1 row, 2 columns of subplots
par(mfrow=c(1,2))
hist(Primates$X25.1_WeaningAge_d)
boxplot(Primates$X25.1_WeaningAge_d)
```

#####Which Order, Primates or Rodentia, has more outliers for InterBirthInterval_d? Do outliers tend to have shorter or longer interbirth intervals? 
```{r}
#Primate outliers; use sort function to put them in order
primate_outliers <- boxplot.stats(Primates$X14.1_InterbirthInterval_d)$out
sort(primate_outliers)
#compare to average
summary(Primates$X14.1_InterbirthInterval_d)
#rodent outliers
rodent_outliers<-  boxplot.stats(Rodentia$X14.1_InterbirthInterval_d)$out
sort(rodent_outliers)
#compare to average
summary(Rodentia$X14.1_InterbirthInterval_d)
```

####Data manipulation with package dplyr
#####grouping and summarizing
```{r}
#1. Group species by Order and Family
by_Order_Family <- group_by(P, MSW05_Order, MSW05_Family)
#2. Summarise to compute means of variables of interest and number of species
by_Ord_Fam_sum <- summarise(by_Order_Family,
  count = n(),
  hr = mean(X22.2_HomeRange_Indiv_km2, na.rm = TRUE),
  mass = mean(X5.1_AdultBodyMass_g, na.rm = TRUE)
)
head(by_Ord_Fam_sum)#look at first 6 rows
```

#####grouping and summarizing using pipe operator
```{r}
#create by_Order_Family to hold summarized info
by_Order_Family_sum <-P %>%
#1. Group species by Order and Family
  group_by(MSW05_Order, MSW05_Family) %>%
#2. Summarise to compute means of variables of interest and number of species
    summarise(
    count = n(),
    hr = mean(X22.2_HomeRange_Indiv_km2, na.rm = TRUE),
    mass = mean(X5.1_AdultBodyMass_g, na.rm = TRUE)
  )
#make a plot of mass vs. hr size with size of dots proportional to number of species
ggplot(data = by_Ord_Fam_sum, mapping = aes(x = mass, y = hr)) +
  geom_point(aes(size = count), alpha = 1/3) +
  geom_smooth(se = FALSE)+
  scale_y_log10()+
  scale_x_log10()
```
